---
title: "Data exploration"
editor_options:
  markdown:
    wrap: 72
---

```{r setup, warning=FALSE, message=FALSE}
library(tidyverse)
library(here)
library(sf)
library(data.table)
library(lubridate)
library(brms)
library(lme4)
```

## Load Q1 data

```{r load Q1 data, warning=FALSE, message=FALSE}
# this datasets contains information of 
trip_changes <- read_csv(here("data", "raw_data", "raw_data_from_king_county", 
                          "Question 1_ Fare Reinstatement", 
                          "alltrips_2020-09_to_2020-10.csv"))

# dates of the trip change are included in the code book and was manually copied here
trip_change_dates <- read_csv(here("data", "manual_data",
                                   "alltrips_change_time.csv"))


apc_routes <- read_csv(here("data", "raw_data", "raw_data_from_king_county", 
                          "Question 1_ Fare Reinstatement", 
                          "apc_detailed_09-01-2020_10-31-2020.csv"))

# The stop level data is the main data set intended to be used 
apc_stops <- fread(here("data", "raw_data", "raw_data_from_king_county", 
                          "Question 1_ Fare Reinstatement", 
                          "stop_activity_granular_2020-09-01_2020-10-31.csv"))

```

## Load SF data

link to ridership data by `STOP_ID` link to ACS data by `GEOID` (derived
using codes in the R script)

```{r load sf data, warning=FALSE, message=FALSE}
kcm_map <- read_sf("data/raw_data/raw_data_from_king_county/KCM_Stops_Data/kcm_stops.shp")
```

## MISC notes for Q1 data

Relevant information 

-   Due to considerable service changes implemented on September 19, 2020, teams may wish to restrict their
analysis to the 9/19/2020 - 10/31/2020 time frame. Of course, teams are welcome to analyze the full time range of provided data (9/01/2020 - 10/31/2020).

-   The information on routes that did not have service changes on
    9/19/20 can be found in the "alltrips" document.

-   Transferring hub stations would have a upward bias for total
    stop-interaction counts, which I (Meng) assume can be identified in
    alltrips document as well.

-   We can discuss the need to subgroup the dates by weekday/weedend
    etc. There is already a column for this in the apc_stops data

+------------------+---------------------------------------------------+
| SCHED_D          | 0 = weekday, 1 = Saturday, 2 = Sunday             |
| A                |                                                   |
| Y_TYPE_CODED_NUM |                                                   |
+==================+===================================================+
+------------------+---------------------------------------------------+

-   KCM confirms there is no way to see when/where a passenger gets on
    and gets off the Metro

-   No data for "Sound Transit", a competitor for KCM, which also
    suspended fare on March 21, 2020, but reinstated fare earlier on
    June 1, 2020.

## data cleanning for Q1

### data cleanning trip changes data

```{r data cleaning for trip changes data}
# only keep the service change code, trip_id and change data to merge with the stop apc data
trip_changes_clean <- trip_changes %>% 
  select(SERVICE_CHANGE_NUM, MINOR_CHANGE_NUM, TRIP_ID) %>% 
  left_join(trip_change_dates, by = c("SERVICE_CHANGE_NUM","MINOR_CHANGE_NUM")) %>%
  rename(change_start_date = `Start Date`,
         change_end_date = `End Date`,
         major_change_code = SERVICE_CHANGE_NUM,
         minor_change_code = MINOR_CHANGE_NUM,
         trip_id = TRIP_ID)
```

### data cleanning trip changes data
I don't think there is any useful information for the trip level data. It has number of passeners boarding and alighting too. But this information can be summarized from the the stop level apc data by `group_by(trip_id)`. There seems to be more operational/administrative information for the trips but we probably won't use those. 

### data cleanning stop level apc data

```{r cleaning apc stops data}
apc_stops_clean <- apc_stops %>% 
# drop BOOKING_ID since we have a stand-alone service change data and drop other variables not to be used
  select(date = OPERATION_DATE,
         day_type = SCHED_DAY_TYPE_CODED_NUM,
         route_id = SERVICE_RTE_LIST,
         stop_id = STOP_ID,
         boarding_num = PSNGR_BOARDINGS,
         alighting_num = PSNGR_ALIGHTINGS,
         trip_id = TRIP_ID) %>% 
  mutate(day_type = case_when(day_type == 0 ~ "weekday",
                              day_type == 1 ~ "Saturday",
                              day_type == 2 ~ "Sunday"))
head(apc_stops_clean)
```

## merge data
```{r merge data}

```





## Explore and data cleaning for stop-level apc data

```{r explore apc_stops, warning=FALSE, message=FALSE}
head(apc_stops)
summary(apc_stops$OPERATION_DATE) # date range 2020-09-01 ~ 2020-10-31

n_trip_id <- apc_stops %>% select(TRIP_ID) %>% 
  count()
n_trip_id_unique <- apc_stops %>% select(TRIP_ID) %>% 
  n_distinct()
n_trip_id - n_trip_id_unique 
# the difference shows that the trip_id is the id for the route, not for the individual passenger
```

```{r derive sub-dataframe for apc_stops, warning=FALSE, message=FALSE}

apc_stops_short <- apc_stops %>% 
# TRIP_ID is kept to link with other two data sets
  group_by(STOP_ID, OPERATION_DATE)%>% 
  summarise(boarding_count = sum(PSNGR_BOARDINGS),
            alighting_count = sum(PSNGR_ALIGHTINGS)) %>% 
mutate(OPERATION_DATE = ymd(OPERATION_DATE)) %>% 
mutate(total_psngr_count = sum(boarding_count + alighting_count),
#KMC suggests using data after 2020/09/19 due to major change on that date but also welcome analyzingthe full time range
       after_major_change = ifelse(OPERATION_DATE >= ymd("2019-09-19"),TRUE, FALSE),
       after_reinstate = ifelse(OPERATION_DATE >= ymd("2020-10-01"), TRUE, FALSE),
         day_number = yday(OPERATION_DATE))

```

```{r run explorative regression, warning=FALSE, message=FALSE}
naive_q1_model <- lm (total_psngr_count ~ after_reinstate, data = apc_stops_short, na.omit = TRUE)
summary(naive_q1_model)
```

```{r model bayes}
model_bayes <- brm(
  bf(total_psngr_count ~ after_reinstate + (1 | STOP_ID) + (1 | day_number)),
  data = apc_stops_short
)

model_bayes
```

```{r model bayes2}
model_bayes <- brm(
  bf(total_psngr_count ~ after_reinstate + (1 | day_number)),
  data = apc_stops_short,
  chains = 4, cores = 4,
)
```

```{r model multilevel}
model_multilevel <- lmer(total_psngr_count ~ after_reinstate + (1 | STOP_ID) + (1 | day_number),
                         data = apc_stops_short)
```

## Model specification plan for Q1

**Question 1:**

`trip_counts ~ after_oct_1_2020 + time_effects + person_effects + stop_effects + route_effects + provider_effects + (fraction bachelors + median_income + weekday) + car ownership and other ACS/census things + pandemic_case_count + national_case_count + subgroups`

## Instructions and Codebook for Q1

Question 1: How did the October 1, 2020 reinstatement of fares affect
ridership on King County Metro? Did the reinstatement of fares by King
County Metro have differential effects on ridership among socio-economic
groups?

- Definitions: By "ridership", we mean the stop-level passenger
boardings (i.e., getting on the bus) and alightings (i.e., getting off
the bus).

- Variables: The data file
stop_activity_granular_2020-09-01_2020-10-31.csv provides automated
passenger counts at the stop-level for Metro buses. The variable
`stop_id` provides the unique identifier for each bus stop. The variable
`psngr_boardings` provides the count of boarding passengers at the
corresponding bus stop. The variable `psngr_alightings` provides the
count of alighting passengers at the corresponding bus stop.

\*\*The stop-level APC data can be linked to the KCM_Stops_Data shape
files using the `stop_id` variable.\* Because passenger counter systems
are not installed on every bus, the ridership data is a sample of
overall system ridership. The data file alltrips_2020-09_to_2020-10.csv
provides schedule information for every trip on the system for each
service change period (typically two weeks or longer). Within each
service change period, trip data is provided for weekday, Saturday, and
Sunday schedules.

+------------------+---------------------------------------------------+
| **Column Name**  | **Description**                                   |
+------------------+---------------------------------------------------+
| OPERATION_DATE   | The date on which the trip was observed in        |
|                  | YYYY-MM-DD format                                 |
+------------------+---------------------------------------------------+
| BOOKING_ID       | Shorthand code for the service change in a format |
|                  | expected by the Hastus NetPlan and ATP modules    |
+------------------+---------------------------------------------------+
| SCHED_D          | 0 = weekday, 1 = Saturday, 2 = Sunday             |
| A                |                                                   |
| Y_TYPE_CODED_NUM |                                                   |
+------------------+---------------------------------------------------+
| SERVICE_RTE_LIST | The customer-facing number of the route (as text  |
|                  | to facilitate the creation of the products based  |
|                  | on this table), i.e. the number used in           |
|                  | timetables, at bus stops, in printed materials,   |
|                  | and on buses' headsigns. This number may be       |
|                  | different than BLOCK_RTE_NUM. All routes have a   |
|                  | service number, even routes that are not known as |
|                  | numbered routes to customers. E.g. A line = route |
|                  | 671.                                              |
+------------------+---------------------------------------------------+
| TR               | Cardinal direction of trip                        |
| I                |                                                   |
| P_COMPASS_DIR_CD |                                                   |
+------------------+---------------------------------------------------+
| KEY_BLOCK_NUM    | Unique identifier of vehicle assignment of trip,  |
|                  | calculated as follows:  (SCHED_DAY_TYPE_CODED_NUM |
|                  | \*100000) + (BLOCK_RTE_NUM \* 100) +              |
|                  | BLOCK_RUN_NUM                                     |
+------------------+---------------------------------------------------+
| S                | Numbered sequence of all bus stops, merging all   |
| TOP_SEQUENCE_NUM | patterns into a single sequence to provide a      |
|                  | proper sequence of stops, regardless of which     |
|                  | pattern a particular trip follows.                |
+------------------+---------------------------------------------------+
| **STOP_ID**      | Unique identifier for each bus stop               |
+------------------+---------------------------------------------------+
| STOP_NM          | The name of the stop                              |
+------------------+---------------------------------------------------+
| SCHED_ARRIVAL\   | The time the bus is scheduled to arrive at the    |
| \_               | stop, expressed as hh:mm:ss without the colons.   |
| SECS_AFTER_MIDNT | Valid only for timestops.                         |
+------------------+---------------------------------------------------+
| FULL_DATE        | The date on which the trip was observed in        |
|                  | MM/DD/YYYY format                                 |
+------------------+---------------------------------------------------+
| ACTUAL_ARRIVAL\  | The time the bus was measured to have arrived at  |
| \_               | the stop, expressed as hh:mm:ss without the       |
| SECS_AFTER_MIDNT | colons                                            |
+------------------+---------------------------------------------------+
| A                | The time the bus was measured to have left the    |
| CTUAL_DEPARTURE\ | stop, expressed as hh:mm:ss without the colons    |
| \_               |                                                   |
| SECS_AFTER_MIDNT |                                                   |
+------------------+---------------------------------------------------+
| PSNGR_BOARDINGS  | The number of passengers that boarded the bus     |
+------------------+---------------------------------------------------+
| PSNGR_ALIGHTINGS | The number of passengers that departed the bus    |
+------------------+---------------------------------------------------+
| **TRIP_ID**      | Trip id                                           |
+------------------+---------------------------------------------------+
| DEP_PSNGR_LOAD   | Number of passengers on the coach                 |
+------------------+---------------------------------------------------+
| Column Name      | Description                                       |
+------------------+---------------------------------------------------+

: Code book for apc_stops




## Load Q2 data

```{r load Q2 data, warning=FALSE, message=FALSE}
lift_boardings <- read_csv("data/raw_data/raw_data_from_king_county/Question 2_ Fare Subsidies/LIFT_boardings.csv")
lift_sales <- read_csv("data/raw_data/raw_data_from_king_county/Question 2_ Fare Subsidies/LIFT_sales.csv")
lift_registry <- read_csv("data/raw_data/raw_data_from_king_county/Question 2_ Fare Subsidies/LIFT_registry.csv")

```

-   Registry means enrollment in the reduced fare program, after
    expiration date, the card become a normal adult card, normally valid
    2 years, but annual pass's registery is good for one year (details
    in May 19 Office Hour record)

-   sales means buying passes, loading money or other "products" to the
    card

-   boardings means using the cards

`card_id` is the shared column



## Explore and analysis - Q2 data

`card_id` is the shared column, we can merge three data sets first by
this column

Re-load counts could be derived from parsing the first several digits (I
have seen 3 digits, 5 digits..."568-251514"), summarize counts. And we
can group by time frames and derive loading counts too.

```{r}
asdf <- lift_registry %>% 
  filter(card_id == "94262-1")
```

## Model specification plan for Q2

**Question 2:**

Indicator for if people reloaded after 6 months

`long_term_loading ~ different_incentives_categorical + age + race + language + issuing_office + home_address_FIPS?`

Indicator for if they have -2 or -3 = renrollment

`reenrollment_in_orca ~ different_incentives_categorical + age + race + language + issuing_office + home_address_FIPS?`

## Instructions and codebook for Q2

### lift registry data 

+---------------+-----------------------------+-----------------------+
| Variable      | Values                      | Description           |
| ID            |                             |                       |
+---------------+-----------------------------+-----------------------+
| Age           | Age at time of enrollment   | Age                   |
|               | (0-101)                     |                       |
+---------------+-----------------------------+-----------------------+
| RaceDesc      | American Indian or Alasa    | Race/Ethnicity        |
|               | Native,                     |                       |
|               | Asian, Black or African     |                       |
|               | American, Hispanic or       |                       |
|               | Latino, Multi-Racial,       |                       |
|               | Native                      |                       |
|               | Hawaiian or Other Pacific   |                       |
|               | Islander, Not Specified,    |                       |
|               | Other, White, NA            |                       |
+---------------+-----------------------------+-----------------------+
| L             | Amharic, Arabic, Cambodian, | Primary language      |
| anguageSpoken | Chinese, Dari, English,     | spoken                |
|               | Farsi, French, Hindi,       |                       |
|               | Japanese, Korean, Laotian,  |                       |
|               | Not                         |                       |
|               | Specified, Oromo, Other,    |                       |
|               | Punjabi, Russian, Samoan,   |                       |
|               | Somali, Spanish, Tagalog,   |                       |
|               | Tigrinya, Ukrainian,        |                       |
|               | Vietnamese, NA              |                       |
+---------------+-----------------------------+-----------------------+
| Expiration    | 4/16/2017 - 9/30/2025 (one  | Date card expires     |
|               | 534984 observation?)        |                       |
+---------------+-----------------------------+-----------------------+
| DateIssued    | 3/1/2017 - 3/17/2022        | Date card was issued\ |
|               |                             | shows date of         |
|               |                             | re-enrollment         |
+---------------+-----------------------------+-----------------------+
| Card          | Various card issuing agency | Agency where card was |
| IssuingAgency | (details in Excel file)     | issued                |
+---------------+-----------------------------+-----------------------+
| FIPS          | 530079602001-530619400023,  | Census block group    |
|               | nan                         | FIPS code of home     |
|               |                             | address (11-digit)    |
+---------------+-----------------------------+-----------------------+
| Initial Load  | 10, 15, 20, 30, 50, 70,     | Product/value         |
|               | Misc.                       | initially loaded onto |
|               | Pass, Monthly Pass,         | card (at no cost to   |
|               | Passport, Subsidized Annual | user)                 |
|               | Pass, NA                    |                       |
+---------------+-----------------------------+-----------------------+
|               |      0                      | Unloaded              |
+---------------+-----------------------------+-----------------------+
|               |      10                     | \$10 stored value     |
+---------------+-----------------------------+-----------------------+
|               |      15                     | \$15 stored value     |
+---------------+-----------------------------+-----------------------+
|               |      20                     | \$20 stored value     |
+---------------+-----------------------------+-----------------------+
|               |      30                     | \$30 stored value     |
+---------------+-----------------------------+-----------------------+
|               |      50                     | \$50 stored value     |
+---------------+-----------------------------+-----------------------+
|               |      Phase1 Passport        | Unlimited ride pass   |
|               |                             | good from roughly     |
|               |                             | Mar/Apr/May/June      |
|               |                             | through Jul/Aug       |
|               |                             | 2019                  |
+---------------+-----------------------------+-----------------------+
|               |      Phase2 Passport        | Unlimited ride pass   |
|               |                             | good from roughly     |
|               |                             | De                    |
|               |                             | c19/Jan20/Feb20/Mar20 |
|               |                             | through Mar           |
|               |                             | 2020                  |
+---------------+-----------------------------+-----------------------+
|               |      Subsidized Annual Pass | Unlimited ride pass   |
|               |                             | good for one year     |
|               |                             | from date of issue    |
+---------------+-----------------------------+-----------------------+
| Study Card    | Phase 1 Treatment, Phase 2  | Shows subsidized      |
|               | Treatment, NA               | annual pass           |
+---------------+-----------------------------+-----------------------+
| duplicate     | TRUE/FALSE (mostly false?)  |                       |
+---------------+-----------------------------+-----------------------+
| card_id       | 100000-1 \~ 99998-3         | Anonymized card ID\   |
|               | (re-enrollment using a dash | - number before the   |
|               | after the card number)      | hyphen indicates      |
|               |                             | user\                 |
|               |                             | - number after hyphen |
|               |                             | indicates card number |
|               |                             | for users who have    |
|               |                             | registered multiple   |
|               |                             | cards                 |
+---------------+-----------------------------+-----------------------+
| Variable      | Values                      | Description           |
| ID            |                             |                       |
+---------------+-----------------------------+-----------------------+

### lift boarding data

+-------------------+-----------------+--------------------------------+
| Variable          | Values          | Description                    |
| ID                |                 |                                |
+-------------------+-----------------+--------------------------------+
| week              | 11/             | Date on which week ends        |
|                   | 7/2021-3/6/2022 |                                |
+-------------------+-----------------+--------------------------------+
| Community Transit | 1\~35, NA       | Boardings made on Community    |
|                   |                 | Transit                        |
+-------------------+-----------------+--------------------------------+
| Everett Transit   | 1-46, NA        | Boardings made on Everett      |
|                   |                 | Transit                        |
+-------------------+-----------------+--------------------------------+
| King County Metro | 1-132, NA       | Boardings made on King County  |
|                   |                 | Metro                          |
+-------------------+-----------------+--------------------------------+
| Kitsap Transit    | 1-23, NA        | Boardings made on Kitsap       |
|                   |                 | Transit                        |
+-------------------+-----------------+--------------------------------+
| Pierce Transit    | 1-40, NA        | Boardings made on Pierce       |
|                   |                 | Transit                        |
+-------------------+-----------------+--------------------------------+
| Sound Transit     | 1-77, NA        | Boardings made on Sound        |
|                   |                 | Transit                        |
+-------------------+-----------------+--------------------------------+
| card_id           | 100042-2 \~     | Anonymized card id; number     |
|                   | 999             | before the hyphen indicates    |
|                   | 98-3(duplicates | user, number after hyphen      |
|                   | by card ID if   | indicates card number          |
|                   | ridden multiple | for users who have registered  |
|                   | weeks)          | multiple cards                 |
+-------------------+-----------------+--------------------------------+
| Variable ID       | Values          | Description                    |
+-------------------+-----------------+--------------------------------+

### lift sales data

+----------------+--------------------------+--------------------------+
| Variable       | Values                   | Description              |
| ID             |                          |                          |
+----------------+--------------------------+--------------------------+
| week           | 3/5/2017-3/6/2022        | Date on which week ends\ |
|                |                          | each enrollee's value    |
|                |                          | loading amount or        |
|                |                          | loading per week         |
+----------------+--------------------------+--------------------------+
| Txn Desc       | Day Pass, Misc. Pass,    | Type of product          |
|                | Monthly                  | purchased                |
|                | Pass, Org. Day Pass,     |                          |
|                | Org. Misc. Pass, Org.    |                          |
|                | Monthly Pass, Org.       |                          |
|                | Purse, Purse             |                          |
+----------------+--------------------------+--------------------------+
|                |      Purse               | "E-purse" stored value   |
+----------------+--------------------------+--------------------------+
|                |      Pass                | Monthly pass product     |
+----------------+--------------------------+--------------------------+
| loadings       | 1-12                     | Number of pass purchases |
|                |                          | or                       |
|                |                          | stored value adds\       |
|                |                          | each enrollee's value    |
|                |                          | loading frequency        |
+----------------+--------------------------+--------------------------+
| Amount         | 0.01-459 (no set         | \$ value of pass         |
|                | increment)               | purschases or            |
|                |                          | stored value adds \      |
|                |                          | each enrollee's value    |
|                |                          | loading amount           |
+----------------+--------------------------+--------------------------+
| card_id        | 100000-1 - 99998-3       |                          |
|                | (duplicates by card ID   |                          |
|                | if ridden                |                          |
|                | multiple weeks)          |                          |
+----------------+--------------------------+--------------------------+

## MISC notes for Q2

The indicators for the RCI treatment and control groups are...
